A Generalized Sphere Theorem for Positively Curved Combinatorial 3-Manifolds

### 2013-04-15 paper/paper.tex
more writing

\begin{thm}[Bonnet-Myers] Let $M$ be a complete, connected, $n$-dimensional Riemannian manifold with section curvature $K \geq \delta > 0$. Then the diameter of $M$ is at most $\frac{\pi}{\sqrt{\delta}}$.
\label{thm:bonnet_myers}
\end{thm}

@@

A very natural discrete definition of distance in an abstract simplicial complex uses \textbf{edge-paths}. That is, paths entirely contained in the 1-skeleton of $M$.

@@

\noindent Just as for edges in an edge-path, the length of each hop and jump will be its length as a PL-path in the standard PL metric. Some basic Euclidean geometry tells us these lengths are $H = \frac{2}{3}\sqrt{2}$ and $J = \frac{1}{2}\sqrt{2} + \sqrt{3}$ respectively. We let $d(v,w)$ denote the \textbf{distance between vertices $v$ and $w$} obtained by minimizing over all paths containing edges as well as hops and jumps. Similarly, we let $diam(M)$ denote the \textbf{diameter of $M$} defined in terms of the distance function $d$.

Why are these paths relevant to this paper? The classical generalized sphere theorem requires the manifold to have diameter more than half the maximum allowed by the classical Bonnet-Myers theorem. If we were to naively imitate the classical results in the discrete setting, from Theorem \ref{thm:discrete_BM} we would want the bound $diam_1(M)>\frac{5}{2}$. Since only integral values of edge-distance can occur, this would correspond to $diam_1(M)\geq 3$. Unfortunately, there are positively curved combinatorial 3-manifolds with edge-diameter three that are not homeomorphic to the 3-sphere. In particular, there exists positively curved triangulations of $\mathbb{RP}^3$ with edge-diameter three. However, if we instead use the finer-grained measure of diameter, $diam(M)$ this problem disappears and the discrete results again mirror the classical ones.

From the proof of Theorem \ref{thm:discrete_BM} in \cite{Trout10}, we have the following bound on the possible distances which occur in a positively curved 3-manifold.

@@

\noindent Note that we have listed the possible distances in increasing numerical order. This result implies a diameter bound, which is known to be sharp by results in \cite{Trout10}.

@@

\noindent This is essentially Theorem \ref{thm:discrete_BM} stated in terms of our new expanded set of paths. It is this result which we will consider our ``discrete Bonnet-Myers'' theorem.

@@

\noindent The proof of this theorem relies on a complete census of positively curved combinatorial 3-manifolds completed by Lutz and Sullivan. The census contains 4787 manifolds, with the following topologies represented: the 3-sphere $S^3$, the real projective space $\mathbb{RP}^3$, the lens spaces $L(3,1)$ and $L(4,1)$ and the cube space $S^3/Q$. The creation of such a census is a highly non-trivial task, for details see \cite{LutzSul}.

@@

This important property is restored if we include hops and edges. In \cite{Trout10} the following is shown.

\begin{thm}
\end{thm}

>>>>

\begin{thm}[Bonnet-Myers] Let $M$ be a complete, connected, $n$-dimensional Riemannian manifold with sectional curvature $K \geq \delta > 0$. Then the diameter of $M$ is at most $\frac{\pi}{\sqrt{\delta}}$.
\label{thm:bonnet_myers}
\end{thm}

@@

A very natural discrete definition of distance in an abstract simplicial complex uses \textbf{edge-paths}. That is, paths entirely contained in the 1-skeleton of $M$. Let $\mathcal{P}_1$ denote the set of all edge-paths on $M$.

@@

\noindent Just as for edges in an edge-path, the length of each hop and jump will be its length as a PL-path in the standard PL metric. Some Euclidean geometry tells us these lengths are \begin{equation} \label{eqn:jump_length} H = \frac{2}{3}\sqrt{2} \end{equation} and \begin{equation} \label{eqn:jump_length} J = \frac{1}{2}\sqrt{2} + \sqrt{3} \end{equation} respectively. We let $d(v,w)$ denote the \textbf{distance between vertices $v$ and $w$} obtained by minimizing over all paths containing edges as well as hops and jumps. Similarly, we let $diam(M)$ denote the \textbf{diameter of $M$} defined in terms of the distance function $d$. Let $\mathcal{P}$ denote the set of all the paths containing edges, hops or jumps.

Why are these paths relevant to this paper? In the classical setting, the generalized sphere theorem requires a manifold to have diameter more than half the maximum allowed by the Bonnet-Myers theorem, and this diameter bound cannot be improved. If we naively imitate the classical results in the discrete setting, from Theorem \ref{thm:discrete_BM} we would want to assume the bound $diam_1(M)>\frac{5}{2}$. Unfortunately, there are positively curved combinatorial 3-manifolds with edge-diameter three that are not homeomorphic to the 3-sphere. In particular, as we will see, $\mathbb{RP}^3$ has positively curved triangulations with edge-diameter three. However, if we instead use the finer-grained measure of diameter, $diam(M)$, the problem disappears and the discrete results mirror the classical ones exactly. That is, we can prove a discrete version of the generalized sphere theorem with a diameter bound exactly half the maximum allowed by the corresponding discrete Bonnet-Myers theorem.

What is the discrete Bonnet-Myers theorem corresponding to Theorem \ref{thm:discrete_BM} but which uses $diam(M)$? From the proof of Theorem \ref{thm:discrete_BM} in \cite{Trout10}, we have the following bound on the possible distances which occur in a positively curved 3-manifold.

@@

\noindent Note that we have listed the possible distances in increasing numerical order. This result implies the desired discrete Bonnet-Myers type theorem.

@@

\noindent Note, it is shown in \cite{Trout10} that this diameter bound is sharp.

@@

\noindent The proof of this theorem relies on a complete census of positively curved combinatorial 3-manifolds completed by Lutz and Sullivan. The census contains 4787 manifolds, with the following topologies represented: the 3-sphere $S^3$, the real projective space $\mathbb{RP}^3$, the lens spaces $L(3,1)$ and $L(4,1)$, and the cube space $S^3/Q$. The creation of such a census is a highly non-trivial task, for details see \cite{LutzSul}.

@@

This important property is restored if we include hops and jumps. In \cite{Trout10} the following is shown.

\begin{thm}(Unique Extension) If two paths $P_1$ and $P_2$ in $\mathcal{P}$ pass through the same first two simplices of $M$ and have the same length, then $P_1 = P_2$. This result is false if $\mathcal{P}$ is replaced by $\mathcal{P}_1$.
\end{thm}
